---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

=========================================================================

## Loading and preprocessing the data
The data is in a `.csv` file named `activity.csv`. Following is the code to load the data into R workspace (assuming that `activity.csv` is present in the workspace), display top 6 rows, and structure of the dataset using `str()`:

```{r}
activity <- read.csv("activity.csv")
head(activity)
str(activity)
```

=========================================================================

## What is mean total number of steps taken per day?
Ignoring missing values in the dataset.

1. Making a histogram of the total number of steps taken each day

```{r}
stepsPerDay <- tapply(activity$steps, activity$date, sum, na.rm=T)
hist(stepsPerDay, main="Histogram of total number of steps per day", xlab="Total steps per day", ylab="Frequency (No. of days)", col="blue")
```

2. Calculating the **mean** and **median** total number of steps taken per day

```{r meanMedian}
stepsMean <- mean(stepsPerDay)
stepsMedian <- median(stepsPerDay)
```

**Mean** --> `r stepsMean`

**Median** --> `r stepsMedian`

=========================================================================

## What is the average daily activity pattern?

1. Make a time series plot of the 5-min interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
activity[,3]<- as.factor(activity$interval)
stepsPerInt <- tapply(activity$steps, activity$interval, mean, na.rm=T)
plot(stepsPerInt, type='l', main="Average Daily Activity Pattern", xlab="5-min interval count", ylab="Number of steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r}
maxStepsInt <- names(which.max(stepsPerInt))
```

The 5-minute interval with maximum average number of steps = `r maxStepsInt`


=========================================================================


## Inputing missing values

1. Total Number of missing values in the dataset.

```{r}
noOfNA <- sum(is.na(activity$steps))
```

Total No of `r NA` --> `r noOfNA`


2. Filling in all the missing values in the dataset. Strategy is to replace NA with 0s

```{r}
steps <- activity$steps
for(i in seq_along(steps)){ 
     if(is.na(steps[i]))
          steps[i]<- 0
}
sum(is.na(steps))
```

3. Creating a new dataset with the new step values without any NAs

```{r}
newActData <- cbind(steps=steps, activity[,c(2,3)])
head(newActData)
str(newActData)
```


4. Histogram of total steps taken each day + mean and median as before on the new data set. Do these values differ from before?

Histogram

```{r}
stepsPerDayNew <- tapply(newActData$steps, newActData$date, sum)
hist(stepsPerDayNew, main="Histogram of total number of steps per day", xlab="Total steps per day", ylab="Frequency (No. of days)", col="blue")
```

Mean and Median


```{r}
stepsMeanNew <- mean(stepsPerDay)
stepsMedianNew <- median(stepsPerDay)
```

**Mean** ---> `r stepsMeanNew`

**Median** ---> `r stepsMedianNew`

There is no change in the result as compared to the first part of the assignment. This shows replacing 0s is not a good strategy for filling up the NA values.


=========================================================================

## Are there differences in activity patterns between weekdays and weekends?

Creating new factor variable with corresponding weekdays.

```{r}
days <- weekdays(as.Date(newActData$date))
for(i in seq_along(days)){ 
     if(days[i] %in% c("Saturday","Sunday"))
          days[i]<- c("weekend")
     else
          days[i]<- c("weekday")
}
days <- as.factor(days)
newActData <- cbind(newActData, days=days)
```

Creating plot

```{r}
weekdaysSubset <- newActData[newActData$days %in% c("weekday"),c(1,3)]
weekendsSubset <- newActData[newActData$days %in% c("weekend"),c(1,3)]
avgStepsWeekdays <- tapply(weekdaysSubset$steps, weekdaysSubset$interval, mean)
avgStepsWeekends <- tapply(weekendsSubset$steps, weekendsSubset$interval, mean)
plot(avgStepsWeekdays, type='l', main="Average Daily Activity Pattern (Weekdays)", xlab="5-min interval count", ylab="Number of steps")
plot(avgStepsWeekends, type='l', main="Average Daily Activity Pattern (Weekends)", xlab="5-min interval count", ylab="Number of steps")


```


=========================================================================